### 2015-02-26 doc/thesis/system.tex
Fixing the section about typing rules to use capitals

In this section we use a reduced core of Typed Lua to present the
most interesting rules of our type system.
This core limits control flow to if and while statements,
has explicit type annotations, and explicit scope for variables.
It also has explicit method declarations and explicit method calls
instead of treating them as syntactic sugar.
We use this reduced core because it simplifies the presentation
of our type system.
Appendix \ref{app:rules} presents the full set of typing rules.

@@

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are literal constants, $p$ are table fields,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

Our reduced core also splits function application, method application, and
the vararg expression (${...}$) into different syntactic categories.
It uses $e(el)_{s}$ to denote function applications that produce no value
because they appear as statements,
$e(el)_{e}$ to denote function applications that produce only one value,
even if the function application returns multiple values,
and $e(el)_{m}$ to denote function applications that produces as many values
as the function application returns.
It uses the same categories for method applications.
For vararg expressions, it only uses ${...}_{e}$ and ${...}_{m}$,
as the vararg expression cannot appear as a statement.

We also include two kinds of type coercions in our core language:
the left-hand value $n[c] \; {<}t{>}$ and the expression ${<}t{>} \;n$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $n_{e}$ when they appear as expressions and $n_{l}$ when they
appear as left-hand values.

We present the typing rules as a deduction system for the typing relations
$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : t, \env_{2}$.
We use the first one for typing statements and the second one for typing expressions.
The first relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a statement $s$ produces
a new type environment $\env_{2}$.
The second relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a expression $e$ has
type $t$ and produces a new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify a type environment.

Even though we can assign only first-level types to variables,
our type system should be able to project unions of first-level types
from unions of second-level types.
Thus, we need two environments because $\env$ maps variables to first-level types,
while $\penv$ maps variables to second-level types.
We will show how our type system uses the environment $\penv$ for
handling projection types.
We use $\env_{1}[n \mapsto t]$ to extend the environment $\env_{1}$
with the variable $n$ that maps to type $t$.
We use $\penv[x \mapsto r]$ to extend the environment $\penv$ with
the variable $x$ that maps to the second-level type $r$.

@@

Lua has multiple assignment, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consistent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

@@

Our type system uses the type $\Value{*}$ to drop extra values
and the type $\Nil{*}$ to replace missing values.
The rule that type checks left-hand values list introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
The rules that type check expression list introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $t{*}$.
In the previous example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
are the rules that introduce these types.
We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:t_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env, \penv \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{i}$.
Then, it takes each type $t_{i}$ and builds the tuple type for
the assignment rule, placing $\Value{*}$ in the end.
Table refinement can change the type environment while typing each left-hand value.
Thus, we use a partial auxiliary function \emph{merge} to collect
all the changes in a new environment $\env_{m}$, if there are no conflicts.

The rule \textsc{T-EXPLIST2} is similar to \textsc{T-LHSLIST}:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env, \penv \vdash e_{i}:t_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env, \penv \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

We use \textsc{T-EXPLIST2} only when the last expression in the list
can produce a single value.
Rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4} handle the case where
an expression list is just an expression that may produce multiple values:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{2}}
      {\env_{1}, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{2}}
\end{array}
\end{array}
\]

As an example, these rules allow our type system to type check the following example:
\[
x, y, z = f()_{m}
\]

@@

Conversely,
\[
x = f()_{m}
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

Rules for function applications are similar to the rule for multiple assignment.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:r_{3}, \env_{3} \;\;\;
       r_{3} \lesssim r_{1}}
      {\env_{1}, \penv \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
or expressions that produce only one value:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e(el)_{s},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e(el)_{e}:first(r), \env_{2}}
\end{array}
\end{array}
\]

The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced.
We can define \emph{first} inductively as follows:
\begin{align*}
first(\Void) & = \Nil\\
first(t{*}) & = t \cup \Nil\\
first(t \times r) & = t\\
first(r_{1} \sqcup r_{2}) & = first(r_{1}) \cup first(r_{2})
\end{align*}

Assuming that $f$ is a local function in the environment, and that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
the function call
\[
f(``foo")_{m}
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the function call
\[
f(``foo",1,2,3)_{m}
\]
also type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

Our type system also catches arity mismatch.
For instance, assuming that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
the function call
\[
f(``foo")_{m}
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but the function call
\[
f(``foo",1,2,3)_{m}
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

@@

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls without optional parameters.
For instance, assuming that $f$ has type
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$,
the function call
\[
f(1)_{m}
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]
and the function call
\[
f(1,2,3)_{m}
\]
also does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]

@@

@@

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of table fields $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions, because we can express it using the first form.
For instance, we can use the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.

The rule \textsc{T-CONSTRUCTOR2} type checks a table type of the first form:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(k_{i},v_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique} \;\;\;
       wf(t)}
      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
\end{array}
\]

First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
$\env_{1}, \penv \vdash p : (k,v), \env_{2}$ to type check each table
field $p_{i}$.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a table field $p$
produces the pair $(k,v)$ and a new type environment $\env_{2}$. 
In the pair $(k,v)$, $k$ is type key type, while $v$ is the value type.

Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(k_{i},v_{i})$
to build the table type that express the type of the given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge k_{i} \lesssim k_{j}
\]

@@

The definition of the auxiliary relation that type checks table
fields would be straightforward if our table types allowed any
first-level type in the type of the keys.
The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD3}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (t_{1},rt(t_{2})), \env_{3}}
\end{array}
\end{array}
\]

These rules use the function \emph{rt} in the type of the value
because our type system allows only \emph{regular} table types to
appear in the type of the values.
The function \emph{rt} promotes all table types to \emph{regular}.
We made this restriction because our type system does not keep track
of aliases to table fields.
This means that allowing \emph{unique}, \emph{open}, and \emph{closed}
table types to appear in the type of the values would allow the creation
of unsafe aliases.

The rule \textsc{T-FIELD4} is a fallback for the previous rules:
\[
\begin{array}{c}
\mylabel{T-FIELD4}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3}}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,rt(t_{2})), \env_{3}}
\end{array}
\]

@@

As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{regular}\}_{unique}$ through the rules
\textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.

Typed Lua handles arrays as hashes that map integers to some type $t$.
In Lua, programmers often use the vararg expression to initialize arrays.
The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
the behavior of the vararg expression for this case:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env({...}) = t}
      {\env, \penv \vdash \{{...}_{m}\}:\{\Integer{:}t \cup \Nil\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env_{i}, \penv \vdash p_{i}:(k_{i},v_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       \env_{i+1}({...}) = t_{v}\\
       t = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}, \Integer{:}t_{v} \cup \Nil\}_{unique} \;\;\;
       wf(t)
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}_{m}\;\}:t, \env_{n+1}}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
only the vararg expression inside a table constructor to initialize
an array.
If we assume that $...$ is in the environment and has type $\String$,
the table constructor $\{{...}_{m}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
through the rule \textsc{T-CONSTRUCTOR3}.

>>>>

In this section we use a reduced core of Typed Lua to present the
most interesting rules of our type system.
Appendix \ref{app:rules} presents the full set of typing rules.

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements are not present because they
are difficult to handle along with our simplified form of
\emph{flow typing}, and they are out of the scope for now;
\item explicit blocks are not present because we are already using
explicit scope for variables;
\item other loop structures such as repeat-until, numeric for,
and generic for are not present because we can use while to
express them;
\item fields in the table constructor are reduced to a single
form because we can use this form to express the missing forms; 
\item arithmetic operators are reduced to addition and unary minus
because the other operators have similar typing rules;
\item relational operators are reduced to equality and less than
because the other operators have similar typing rules;
\item bitwise operators are not present because they are similar
to integer arithmetic operators.
\end{itemize}

@@

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $k$ are literal constants, $p$ are table fields,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $a$ are function and method applications,
$n$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\vec{n{:}T}$ denotes the list $n_{1}{:}T_{1}, ..., n_{n}{:}T_{n}$,
where $n$ can be zero.

Our reduced core includes two statements for declaring local variables.
One with explicit type annotations, and another without type annotations.
While we use the former to show how our type system handles the declaration
of annoatted variables, we use the latter to show how our type system
handles the declaration of unannotated variables through local type inference
and also the introduction of projection types.

Our reduced core also includes a truncation operator $\lfloor \rfloor$ for
function applications, method applications, and the vararg expression.
We use $\lfloor a \rfloor_{0}$ to denote function and method applications
that produce no value because they appear as statements.
We use $\lfloor a \rfloor_{1}$ to denote function and method applications
that produce only one value, even if the function or method application
returns mutiple values.
We use $\lfloor {...} \rfloor_{1}$ to denote a vararg expression that
produces only one value, as it can appear in the middle of an
expression list.

We also include two kinds of type coercions in our core language:
the left-hand value $n[k] \; {<}T{>}$ and the expression ${<}T{>} \;n$.
Both allow the refinement of table types.
We also split variable names into two categories to have safe aliasing
of tables in the presence of refinement.
We use $n$ when they appear as expressions and $n_{l}$ when they
appear as left-hand values.

We present the typing rules as a deduction system for the typing relations
$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : T, \env_{2}$.
We use the first one for typing statements and the second one for typing expressions.
The first relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a statement $s$ produces
a new type environment $\env_{2}$.
The second relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing an expression $e$ has
type $T$ and produces a new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify a type environment.

Even though we can assign only first-level types to variables,
our type system should be able to project unions of first-level types
from unions of second-level types.
Thus, we need two environments because $\env$ maps variables to first-level types,
while $\penv$ maps variables to second-level types.
We will show how our type system uses the environment $\penv$ for
handling projection types.
We use $\env_{1}[n \mapsto T]$ to extend the environment $\env_{1}$
with the variable $n$ that maps to type $T$.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
the variable $x$ that maps to the second-level type $S$.

@@

Lua has multiple assignment, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consistent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

@@

Our type system uses the type $\Value{*}$ to drop extra values
and the type $\Nil{*}$ to replace missing values.
The rule that type checks left-hand values list introduces the type
$\Value{*}$ to let the right-hand side produce more values than
expected in the left-hand side.
The rules that type check expression list introduce the type
$\Nil{*}$ to let the right-hand side produce fewer values than
expected in the left-hand side.
Our type system includes $\Nil{*}$ in the type of an expression list only
if its type does not end in another variadic type $t{*}$.
In the previous example, the rules \textsc{T-LHSLIST} and \textsc{T-EXPLIST2}
are the rules that introduce these types.
We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env, \penv \vdash \vec{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{i}$.
Then, it takes each type $T_{i}$ and builds the tuple type for
the assignment rule, placing $\Value{*}$ in the end.
Table refinement can change the type environment while typing each left-hand value.
Thus, we use a partial auxiliary function \emph{merge} to collect
all the changes in a new environment $\env_{m}$, if there are no conflicts.

The rule \textsc{T-EXPLIST2} is similar to \textsc{T-LHSLIST}:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env, \penv \vdash \vec{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

We use \textsc{T-EXPLIST2} only when the last expression in the list
can produce a single value.
Rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4} handle the case where
an expression list is just an expression that may produce multiple values:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Void, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
      {\env_{1}, \penv \vdash me:T_{1} \times ... \times T_{n}{*}, \env_{2}}
\end{array}
\end{array}
\]

As an example, these rules allow our type system to type check the following example:
\[
x, y, z = f()
\]

@@

Conversely,
\[
x = f()
\]
type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \Value{*}
\]

Rules for function applications are similar to the rule for multiple assignment.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that produce multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:S_{1} \rightarrow S_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:S_{3}, \env_{3} \;\;\;
       S_{3} \lesssim S_{1}}
      {\env_{1}, \penv \vdash e(el):S_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
or expressions that produce only one value:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{0},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el):S, \env_{2}}
      {\env_{1}, \penv \vdash \lfloor e(el) \rfloor_{1}:first(S), \env_{2}}
\end{array}
\end{array}
\]

The rule \textsc{T-STMAPPLY1} discards the produced values,
while the rule \textsc{T-EXPAPPLY1} uses the auxiliary function
\emph{first} to ensure that only one value is produced.
We can define \emph{first} inductively as follows:
\begin{align*}
first(\Void) & = \Nil\\
first(T{*}) & = T \cup \Nil\\
first(T \times P) & = T\\
first(S_{1} \sqcup S_{2}) & = first(S_{1}) \cup first(S_{2})
\end{align*}

Assuming that $f$ is a local function in the environment, and that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
the function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the function call
\[
f(``foo",1,2,3)
\]
also type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

Our type system also catches arity mismatch.
For instance, assuming that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
the function call
\[
f(``foo")
\]
type checks through the rule \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but the function call
\[
f(``foo",1,2,3)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

@@

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls without optional parameters.
For instance, assuming that $f$ has type
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$,
the function call
\[
f(1)
\]
does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]
and the function call
\[
f(1,2,3)
\]
also does not type check through the rule \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]

@@

Our abstract syntax reduces the syntactic forms of the table constructor
into two forms: $\{\;\vec{p}\;\}$ and $\{\;\vec{p},me\;\}$.
The first uses a list of table fields $p_{1}, ..., p_{n}$,
where $p$ is $[e_{1}] = e_{2}$ and $n$ can be zero.
The second uses a list of table fields and an expression that can
produce multiple values.
We did not include in the abstract syntax a table constructor that uses
lists of expressions, because we can express this form using lists of
table fields.
For instance, we can use the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.

@@

The rule \textsc{T-CONSTRUCTOR2} type checks a table type with a non-empty
list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T)}
      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:T, \env_{n+1}}
\end{array}
\]

First, the rule \textsc{T-CONSTRUCTOR2} uses the auxiliary relation
$\env_{1}, \penv \vdash p : (K,V), \env_{2}$ to type check each table
field $p_{i}$.
This auxiliary relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, typing a table field $p$
produces the pair $(K,V)$ and a new type environment $\env_{2}$. 
In the pair $(K,V)$, $K$ is the key type, while $V$ is the value type.

Then, the rule \textsc{T-CONSTRUCTOR2} uses each pair $(K_{i},V_{i})$
to build the table type that express the type of the given constructor, and
uses the predicate \emph{wf} to check whether this table type is well-formed.
Formally, a table type is well-formed if it obeys the following rule:
\[
\forall i \not\exists j \; i \not= j \wedge K_{i} \lesssim K_{j}
\]

@@

The definition of the auxiliary relation that type checks table
fields would be straightforward if our table types allowed any
first-level type in the type of the keys.
The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
       T_{1} \subtype \Boolean}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
       T_{1} \subtype \Number}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD3}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3} \;\;\;
       T_{1} \subtype \String}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (T_{1},close(T_{2})), \env_{3}}
\end{array}
\end{array}
\]

These rules use the function \emph{close} in the type of the value
because our type system allows only \emph{closed} table types to
appear in the type of the values.
The function \emph{close} promotes all table types to \emph{closed}.
We made this restriction because our type system does not keep track
of aliases to table fields.
This means that allowing \emph{unique}, \emph{open}, and \emph{fixed}
table types to appear in the type of the values would allow the creation
of unsafe aliases.

The rule \textsc{T-FIELD4} is a fallback for the previous rules:
\[
\begin{array}{c}
\mylabel{T-FIELD4}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:T_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:T_{2}, \env_{3}}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: (\Value,close(T_{2})), \env_{3}}
\end{array}
\]

@@

As an example, the table constructor $\{[``x"] = 1, [``y"] = \{[``z"] = 2\}\}$
has type $\{``x":1, ``y":\{``z":2\}_{closed}\}_{unique}$ through the rules
\textsc{T-FIELD3} and \textsc{T-CONSTRUCTOR2}.

Typed Lua handles arrays as hashes that map integers to some type $T$.
In Lua, programmers often use the vararg expression to initialize arrays.
The rules \textsc{T-CONSTRUCTOR3} and \textsc{T-CONSTRUCTOR4} define
the behavior of the vararg expression for this case:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-CONSTRUCTOR3}\\
\dfrac{\env({...}) = T}
      {\env, \penv \vdash \{{...}\}:\{\Integer{:}T \cup \Nil\}_{unique}, \env}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-CONSTRUCTOR4}\\
\dfrac{\begin{array}{c}
       \env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       \env_{i+1}({...}) = T_{v}\\
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, \Integer{:}T_{v} \cup \Nil\}_{unique} \;\;\;
       wf(T)
       \end{array}}
      {\env_{1}, \penv \vdash \{\;\vec{p},\;{...}\;\}:T, \env_{n+1}}
\end{array}
\end{array}
\]

The rule \textsc{T-CONSTRUCTOR3} type checks the case where we use
only the vararg expression inside a table constructor to initialize
an array.
If we assume that $...$ is in the environment and has type $\String$,
the table constructor $\{{...}\}$ has type $\{\Integer:\String \cup \Nil\}_{unique}$
through the rule \textsc{T-CONSTRUCTOR3}.

The rule \textsc{T-CONSTRUCTOR4} type checks the case where we use
the table constructor as a record that includes an array part.
If we assume that $...$ is in the environment and has type $\String$,
the table constructor $\{[``x"] = ``foo", [``y"] = \mathbf{true}, {...}\}$
has type $\{``x":``foo", ``y":\mathbf{true}, \Integer:\String \cup \Nil\}_{unique}$
through the rule \textsc{T-CONSTRUCTOR4}.
As another example, the rule \textsc{T-CONSTRUCTOR4} does not type check
the table constructor $\{ [1] = ``foo", [10] = \mathbf{true}, {...} \}$,
because the keys $1$ and $10$ overlap the array part,
and the resulting table type is not well-formed.
